Real-time Attention Assessment for Social Robots

### 2015-09-16 paper.tex
Work on literature + technique + added/updated figures

The relations between one's focus of attention and what he/she is looking at has
long been established~\cite{yarbus1967eye,barber1976perception}, and more
specifically, the existing relationship between gaze and attention during social
interaction and the related gaze patterns, belongs to classic textbooks
like~\cite{argyle1969social}. As such, measuring the direction of gaze is
a natural proxy to estimate the (visual) focus of attention of a social agent.

@@

\subsection{Overview of the Technique}

We propose a adaptive model based solution to approach head pose estimation that
finds correspondences between facial landmarks points and their respective
locations in a face model. We use the open-source library
dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton
of the subject's face and then, use the obtained features to calculate the
head pose orientation estimation (see figure \ref{system}). In this way, we can
estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
representation as it is executed as a ROS (Robot Operating System) node that can
be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted
displays, eye-trackers or wearables for instance, this approach has a clear
advantage of simplicity and integration with robotic systems, but also becomes
accurate enough for general purposes in face-to-face interactions. In addition,
it allows us to model complex environments, handling different number and
positions of participants in the scene.

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{rviz_camera}
    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFoA.}
    \label{rviz}
\end{figure}

\subsection{Estimating Head Pose}
The first step is to estimate the head pose using the pin-hole model shown in
equation \ref{eq:pinHole}. As mentioned above, dlib library facilitates the task
by providing the 2D points \textit{m'} that composes the face silhouette.
However, it is enough with obtaining eight spread points, such as right and left
eyes, right and left tragions, sellion or nasal depression zone, pronasale or
center of the nose, stomion or center of the mouth and menton.

Likewise, the 3D face model is needed. The proposed head pose estimation method
would work best with a 3D face model of the query subject itself. When a 3D face
model of the subject is not available, as in many practical situations, a
generic 3D face model can be used. This is an estimation of the same
anthropometric points \textit{M'} (for North American Caucasians) in the 3D
space \cite{farkas1994anthropometry}. Moreover, the intrinsic camera parameters
are necessary to be able to compose its matrix \textit{A}. The components are
expanded in equation \ref{eq:pinHole2}.

\begin{equation}
s  \; m' = A [R|t] M'
\label{eq:pinHole}
\end{equation}
\begin{equation}
s
\begin{bmatrix}
x \\
y \\
1
\end{bmatrix}
=
\begin{bmatrix}
f_x & 0 & c_x  \\
0 & f_y & c_y  \\
0 & 0 & 1 
\end{bmatrix}
\begin{bmatrix}
r_{11} & r_{12} & r_{13} & t_x  \\
r_{21} & r_{22} & r_{23} & t_y  \\
r_{31} & r_{32} & r_{33} & t_z  
\end{bmatrix}
\begin{bmatrix}
X \\
Y \\
Z \\
1
\end{bmatrix}
\label{eq:pinHole2}
\end{equation}

Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the
rotation matrix R (calculated from the rotation vector) and translation vector
\textit{t}. The result is a matrix projection \textbf{P} which maps a point in
the 3D space onto a point in the 2D image plane solving equation \ref{eq:p}.

\begin{equation}
s  \; m' = P \; M'
\label{eq:p}
\end{equation}

\subsection{Acquiring the View of Attention}

@@

Defining accurately the human field of view in the implementation is a must.
Holmqvist in \cite{holmqvist2011eye} specifies that the visual human range is $
\pm  40\degree $ in the horizontal and $ \pm 25\degree $ in the vertical. In
\cite{walker1980clinical} Spector provides a more detailed specification for
each eye splitting the vertical range into $ 60\degree $ the upper region and $
75\degree $ the lower one. Moreover, the horizontal range gets separated in $
60\degree $ inwards (towards the nose) and $ 95\degree $ outwards. In the
present implementation, the first approach explained has been chosen
representing the field of view using a cone with such dimensions.

@@

>>>>

The relations between one's focus of attention and what he/she is looking at has
long been established~\cite{yarbus1967eye,barber1976perception}, and more
specifically, the existing relationship between gaze and attention during social
interaction, and the related gaze patterns, has been part of classic textbooks
like~\cite{argyle1969social} since decades. As such, there is little doubt that
measuring the direction of gaze is a useful proxy to estimate the (visual) focus
of attention of a social agent, and indeed this is one of the basic tool used in
social psychology.

Estimating attention using gaze is certainly not new to robotics either. A
recent survey by Ruhland~\etal~\cite{ruhland2015review} presents in a synthetic
way eye gaze research in HCI and social robotics, and is an essential read in
that regard. It remains however an active field of research, as illustrated by
several recent publications~\cite{baxter2014tracking,anzalone,kennedy2015head}:
as attention tracking is essential in human-robot interaction

@@

\subsection{Head Pose Estimation}

As explained, we derive the visual field of attention from the head pose. Our
technique only involves a single monocular RGB camera for used for facial
feature extraction, and a static simplified 3D mesh of a human head
(Figure~\ref{head_pose}). 68 facial features are extracted using a fast
template-based face alignment algorithm by Kazemi and
Sullivan~\cite{kazemi2014one}, as implemented in the open-source {\tt dlib}
library\footnote{http://dlib.net}~\cite{dlib09}.  Eight of these features are
matched to their 3D counterparts and we rely on an iterative $PnP$ algorithm
(OpenCV's implementation) to compute the translation and rotation of the head
with respect to the camera frame. With this approach, knowing the intrinsic
parameters of the camera (calibrated camera) is required for an accurate
estimation of the absolute 3D localization of the head.

Besides being fast, the face alignment algorithm has been found to perform well,
including in a range of difficult situations (occlusions, head angle) as
encountered in field experiments.  Figure~\ref{head_pose_real_world} shows a
selection of such difficult scenes on one child.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\linewidth]{head_pose}
    \caption{The 6D head pose is estimated by fitting a 3D model of an
        adult head (left) onto the detected 2D features of the face (right). We
        rely on an iterative $PnP$ algorithm, using 8 correspondence pairs
        (three are depicted: the sellion -- the nasal depression --, the left
        tragion and the menton).}

    \label{head_pose}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{head_pose_real_world}
    \caption{Head pose results on images captured during a field experiment.
    Detection of face features (and therefore, estimation of the pose) is robust
    to significant occlusions and face rotations.}
    \label{head_pose_real_world}
\end{figure}

\subsection{Field of Attention}

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{field_of_attention}
    \caption{\small The visual field of attention is approximated to a
        20$\degree$ cone, spanning from the head's sellion. The objects whose 3D
    pose intersect with this cone are considered \emph{in focus}.}
    \label{rviz}
\end{figure}

, to detect and extract the skeleton
of the subject's face and then, use the obtained features to calculate the
head pose orientation estimation (see figure \ref{system}). In this way, we can
estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
representation as it is executed as a ROS (Robot Operating System) node that can
be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted
displays, eye-trackers or wearables for instance, this approach has a clear
advantage of simplicity and integration with robotic systems, but also becomes
accurate enough for general purposes in face-to-face interactions. In addition,
it allows us to model complex environments, handling different number and
positions of participants in the scene.

@@

The breadth of the human field of attention has been discussed in the
literature.
Holmqvist in~\cite{holmqvist2011eye} specifies that the visual human
range\fixme{field of view or field of attention?} is $
\pm  40\degree $ in the horizontal and $ \pm 25\degree $ in the vertical. In
\cite{walker1980clinical} Spector provides a more detailed specification for
each eye splitting the vertical range into $ 60\degree $ the upper region and $
75\degree $ the lower one. Moreover, the horizontal range gets separated in $
60\degree $ inwards (towards the nose) and $ 95\degree $ outwards. In robotics,
Sisbot\etal~\cite{sisbot2011situation} use a cone of 30$\degree$. In the
present implementation, the first approach explained has been chosen
representing the field of view using a cone with such dimensions\fixme{which
one? 40deg? 25deg?}.

@@

\subsection{System Implementation}
